Supplementary Report for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*

### 2023-01-25 reports/supplement/supplement_0.9.1.Rmd
Update supplement_0.9.1.Rmd

@@

We show in the main text that diversity in the sulfate-reducing bacteria results in earlier recovery of the anoxic state (Figure 3 c). However, there is only limited evidence of strain replacement on the trajectory of decreasing oxygen diffusivity: the most tolerant strain does increase in density prior to the tipping point but never reaches dominance – it is the strain with lowest tolerance that dominates immediately after the shift to the anoxic state. We here analyze temporal dynamics rather than the final state and show that the more tolerant strains do play a role, but only a transient one. The following dynamics are for the system starting oxic, and with a value of oxygen diffusivity $log_{10}$(oxygen diffusivity) = -4.8. When there is no diversity within functional groups (i.e. two strains that are identical) the system remains oxic (Figure \@ref(fig:sb-div-effect3)). When there is diversity in the two groups of sulfur bacteria (i.e. the two strains differ) the system switches to the anoxic state (Figure \@ref(fig:sb-div-effect4)). The more tolerant strain of sulfate-reducing bacteria increases in density in the early phase of the simulation, reduces the concentration of oxygen by inhibiting cyanobacteria, and thus generates an environment in which the less tolerant strain is able to grow. Then, the more tolerant strain is outcompeted by the less tolerant strain as the concentration of the inhibiting substrate (oxygen) declines.

>>>>

We have confirmed that when there is no within functional group strain variation, the dynamics of the model are not affected by the number of strains in each of the functional groups. This is a necessary condition for such a model [Moisset de Espanés et al. 2021](https://onlinelibrary.wiley.com/doi/10.1111/ele.13775), and gives confidence that our methods and their implementation are correct and robust.

@@

We show in the main text that diversity in the sulfate-reducing bacteria results in earlier recovery of the anoxic state (Figure 3 c). However, there is only limited evidence of strain replacement on the trajectory of decreasing oxygen diffusivity: the most tolerant strain does increase in density prior to the tipping point but never reaches dominance – it is the strain with lowest tolerance that dominates immediately after the shift to the anoxic state. We here analyze temporal dynamics rather than the final state and show that the more tolerant strains do play a role, but only a transient one. The following dynamics are for oxic starting conditions and with a value of oxygen diffusivity $log_{10}$(oxygen diffusivity) = -4.8. When there is no diversity within functional groups (i.e. two strains that are identical) the system remains oxic (Figure \@ref(fig:sb-div-effect3)). When there is diversity in the two groups of sulfur bacteria the system switches to the anoxic state (Figure \@ref(fig:sb-div-effect4)). (Note that the average trait value [on a log scale] of the two different strains is the same as the trait value of both strains in Figure \@ref(fig:sb-div-effect3).) The more tolerant strain of sulfate-reducing bacteria increases in density in the early phase of the simulation, reduces the concentration of oxygen by inhibiting cyanobacteria, and thus generates an environment in which the less tolerant strain is able to grow. Then, the more tolerant strain is outcompeted by the less tolerant strain as the concentration of the inhibiting substrate (oxygen) declines.
